---
title: "Spatial Analysis with Movement Patterns at Finer Scales"
author: "Prince Oppong Boakye"
date: "2025-05-15"
output: html_document
---

```{r setup, include=FALSE}
  knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE)
```

## Setup (Mapbox Token)

```{r}
Sys.setenv(MAPBOX_TOKEN = "ADD YOUR MAPBOX TOKEN")
```

## Libraries

```{r}
library(spanishoddata)
library(flowmapblue)
library(tidyverse)
library(sf)
library(dplyr)
library(ggplot2)
```

## Data Loading & Processing

```{r}
setwd("~/Desktop/SDSWR Assignment 1/dataanalysis")

spod_set_data_dir(data_dir = "~/dataanalysis/spanish_od_data")

```

## Analysis

```{r}
od_20210407 <- spod_get("od", zones = "distr", dates = "2021-04-07")

head(od_20210407)
```

```{r}
districts_v1 <- spod_get_zones("dist", ver = 1)

head(districts_v1)
```

```{r}
str(flowmapblue::ch_locations)
str(flowmapblue::ch_flows)
```

```{r}
od_20210407_total <- od_20210407 |>
  group_by(origin = id_origin, dest = id_destination) |>
  summarise(count = sum(n_trips, na.rm = TRUE), .groups = "drop") |> 
  collect()

head(od_20210407_total)
```

```{r}
districts_v1_centroids <- districts_v1 |>
  st_transform(4326) |> 
  st_centroid() |>
  st_coordinates() |>
  as.data.frame() |>
  mutate(id = districts_v1$id) |>
  rename(lon = X, lat = Y)

head(districts_v1_centroids)
```

### Flowmap (Spain-Clustered)

```{r}
flowmap <- flowmapblue(
  locations = districts_v1_centroids,
  flows = od_20210407_total,
  mapboxAccessToken = Sys.getenv("MAPBOX_TOKEN"),
  darkMode = TRUE,
  animation = FALSE,
  clustering = TRUE
  )

flowmap
```

### Flowmap (Date & Time Series)

```{r}
od_20210407_time <- od_20210407 |>
  mutate(time = as.POSIXct(paste0(date, "T", hour, ":00:00"))) |>
  group_by(origin = id_origin, dest = id_destination, time) |>
  summarise(count = sum(n_trips, na.rm = TRUE), .groups = "drop") |> 
  collect()

head(od_20210407_time)
```

### Flowmap (Spain-Non-Clustered)

```{r}
spain_fineScale <- flowmapblue(
  locations = districts_v1_centroids,
  flows = od_20210407_total,
  mapboxAccessToken = Sys.getenv("MAPBOX_TOKEN"),
  clustering = FALSE,
  darkMode = TRUE,
  animation = FALSE
)
spain_fineScale
```

### Flowmap + Date & Time Series)

```{r}
spain_fineScaleTimeSeries <- flowmapblue(
  locations = districts_v1_centroids,
  flows = od_20210407_time,
  mapboxAccessToken = Sys.getenv("MAPBOX_TOKEN"),
  clustering = FALSE,
  darkMode = TRUE,
  animation = FALSE
)
spain_fineScaleTimeSeries
```

### Madrid Districts

```{r}
zones_madrid <- districts_v1 |>
  filter(grepl("Madrid distrito", district_names_in_v2, ignore.case = TRUE))

zones_madrid_fua <- districts_v1[
  st_buffer(zones_madrid, dist = 10000)
  ,
]

zones_madrid_fua_plot <- ggplot() +
  geom_sf(data = zones_madrid_fua, fill=NA, col = "grey60", linewidth = 0.3) +
  theme_minimal()

zones_madrid_fua_plot
```

```{r}
zones_madrid_fua_coords <- zones_madrid_fua |>
  st_transform(crs = 4326) |>
  st_centroid() |>
  st_coordinates() |>
  as.data.frame() |>
  mutate(id = zones_madrid_fua$id) |>
  rename(lon = X, lat = Y)

head(zones_madrid_fua_coords)
```

### Flowmap (Madrid-Clustered)

```{r}
od_20210407_time_madrid <- od_20210407_time |>
  filter(origin %in% zones_madrid_fua$id & dest %in% zones_madrid_fua$id)

madrid_fineScaleTimeSeries <- flowmapblue(
  locations = zones_madrid_fua_coords,
  flows = od_20210407_total,
  mapboxAccessToken = Sys.getenv("MAPBOX_TOKEN"),
  darkMode = TRUE,
  animation = FALSE,
  clustering = FALSE
  )

madrid_fineScaleTimeSeries
```

### Flowmap (Madrid-Non-Clustered + Date & Time Series)

```{r}
madrid_fineScaleTimeSeries <- flowmapblue(
  locations = zones_madrid_fua_coords,
  flows = od_20210407_time_madrid,
  mapboxAccessToken = Sys.getenv("MAPBOX_TOKEN"),
  darkMode = TRUE,
  animation = FALSE,
  clustering = FALSE
  )

madrid_fineScaleTimeSeries
```

### Total flows Madrid District + 10Km FUA Heatmap

```{r}
out_flows <- od_20210407_total |>
  group_by(id = origin) |>
  summarise(out_flows = sum(count))

in_flows <- od_20210407_total |>
  group_by(id = dest) |>
  summarise(in_flows = sum(count))

flow_totals <- out_flows |>
  full_join(in_flows, by = "id") |>
  mutate(
    out_flows = replace_na(out_flows, 0),
    in_flows = replace_na(in_flows, 0),
    total_flows = out_flows + in_flows
  )

madrid_activity <- zones_madrid_fua |>
  left_join(flow_totals, by = "id") |>
  st_as_sf()


ggplot(madrid_activity) +
  geom_sf(aes(fill = total_flows), color = "grey", alpha = 0.8) +
  scale_fill_gradientn(colors = RColorBrewer::brewer.pal(9, "YlOrRd")) +
  labs(title = "Madrid Mobility Flows") +
  theme_minimal()
```

## Research Questions

1.  How do the mobility patterns in Madrid differ from those in other regions of Spain?
2.  How do mobility-based contact points (created from simulated flows) relate to historical patterns of disease spread, and in restricted movement scenarios can that reduce hypothetical transmissions?
3.  To what extent could we simulate fine-scale population flows by harnessing multi-source datasets to approximate realistic contact patterns for disease transmission models in Madrid?
